Parallelization of Push-based System for Molecular Simulation Data Analysis with GPU

### 2016-02-25 paper/paperSource.tex
automatic triggered changes

\emph{\textbf{SDH implementation.}} Even though implementation of SDH can be different, in the previous paper the algorithm in the previous paper was based on quad tree data structure with data proximity approach based on data uniformity\cite{mainPaper}, but in our experiments we follow special CUDA based approach which perfectly fits the nature of GPU devices with the given memory hierarchy, and the number of computations are equal to brute force, consequently, there is no expectation on data uniformity, thus, the problem solution complexity is still the same with the worst case of $O(N^2)$, but processed in parallel by GPU cores.

The following set of one-body
queries were included in the test workload: mean square displacement
(msd), radius of gyration, dipole moment, center of mass, velocity
autocorrelation, electron density, mass density, and charge density.
This set of queries were pointed to us, by a group in the physics
field with extensive MS background, as one of the most commonly used
in the field of collagen bilayer MS system analysis. A workload
group contains all $8$ queries executed on one of the $12$
selections, making $12$ groups. Such groups are executed on six
different size data sets, with $10$, $100$ and $1000$ frames. This
workload is then repeated $5$ more times, by executing each of the
queries in the groups $5$, $10$, $25$, $50$, and $100$ times,
essentially just magnifying the workload intensity. In total, we
have $12 x 6 x 3 x 6 = 1,296$ different workload setups to test the
system on. Fig.~\ref{fg:workload_set_up} shows the organization of
the workload setup.

@@

\subsection{Benchmark}
Through extensive collaboration with a research group from the
Physics department at USF, we have come up with a benchmark that can
be used for testing the efficiency of an analysis system for
molecular simulations. The benchmark consist of three essential
parts: 1. Simulation data produced by an MS, 2. Queries that are to
be executed onto that data in order to produce some information of
interest, and 3.Benchmark parameters that control the size of the
benchmark.

\subsubsection{Benchmark Data}
The data used in the benchmark was real molecular simulation data,
produced through the GROMACS MS system. The initial, pre-simulation
data file consisted of $200$ POPC and $12000$ solvent molecules, or
$12200$ molecules in total. This type of system was used because it
is sufficiently diverse, containing enough distinct POPC and solvent
molecules (e.g., each POPC molecule includes approximately $52$
different atoms) and yet simple enough to be easily transformed into
another system of different size. By using the $genconf$ function in
GROMACS, we produced pre-simulation files of different sizes
(essentially by changing the system's size (box)). $Six$ different
sized pre-simulation files were created. A molecular simulation was
then run on these $6$ files, each producing an MS system of certain
size (volume/number of particles). All of the simulations were set
up to produce $1000$ frames (snapshots in time of the systems), each
frame containing the same number of particles as the base one. The
produced files contained: $52,400$; $209,600$; $838,400$; $2.5M$;
$4.4M$; and $8.8M$ atoms per frame. So, for example, the file with
$52,000$ atoms holds $52,000,000$ records in total ($1000$ frames,
each containing $52,000$ records). As mentioned earlier, this
simulation data comes mostly in binary formats and in trajectory
files having a lot of unneeded overhead. Therefore, it was
transformed to a data arrays files containing only crucial
information of the particles and the system. The size of the files
ranged from $135MB$ for $52,000$ atoms to $24GB$ for $8.8$ million
atoms (this is for data with $100$ frames).

%Fig.~\ref{fg:dataformat} represents the organization/structure of
%these files.

\subsubsection{Benchmark Queries}
The queries selected to be included in this benchmark were derived
through a thorough observation of the way an MS system is being
analyzed. They were found to be the base of the analysis of many MS
systems. In other words, no mater how small or big the analysis was,
these queries were included in that analysis. As mentioned earlier,
they are of two types: one-body (and algebraic) and two-boy (and
holistic). Table~\ref{tb:queries} shows these queries.

\subsubsection{Benchmark Parameters} There are several parameters
that can be used to control the overall size of the system. We divide the parameters into two groups:\\
Data size parameters:
\begin{itemize}
\item[-]Select different sized dataset
\item[-]Number of frames
\item[-]Data selection (within the selected dataset) onto which the queries are being executed
\end{itemize}
Workload size parameters:
\begin{itemize}
\item[-]Number of queries to be executed
\item[-]Number of times each query is executed
\end{itemize}
By changing these parameters, we can produce a versatile testing
benchmark for MS analysis systems.

\subsection{Results} We have run extensive experiments over all the
different setups of workload mentioned previously in this section.
However, in this paper we present only the workload setups of $4$
data size sets: $838,400$, $2.5M$, $4.2M$, and $8.8M$ atoms because
we believe they convey enough information about the efficiency of
our system compared to that of the Gromacs system. The running times
of our push-based system were compared to those of the Gromacs
system. The first set of figures, namely
Figures~\ref{fg:speedup838K-select-level}-\ref{fg:speedup8.8M-select-level},
represent the speedup that our system obtains over the Gromacs
system with various atoms selection levels. We define the selection
levels based on the number of comparisons we have to make in order
to extract the needed group(selection) of atoms. For example, if we
want to do analysis on all molecules containing oxygen, or hydrogen,
or carbon we would go over each molecule and compare its components
to the selection list. The bigger the selection list, the higher the
select level in our system. For better visualization, we note three
different selection levels: high (at least $10$ comparisons made),
medium (between $1$ and $10$ comparisons made), and low select level
(with one or less comparisons made). As seen in the figures, for
high selection level, the speedup is smaller compared to that
achieved in low select levels. The reason for this, we believe is in
that the amount of time our system spends extracting the atoms group
increases with the level of selection. Even though our system still
shows considerable speedup over Gromacs in high level selections, we
do believe there is room for improvement in our system and that is
our immediate future work we are planning on doing. These figures
also show the relation of the speedup to the workload intensity,
i.e., the higher the workload intensity the higher the speedup.

The connection between the workload intensity and the speedup is
better represented in the next set of figures,
Figures~\ref{fg:speedup838K}-\ref{fg:speedup8.8M}. They show the
speedup our system achieves over the Gromacs system on a varying
workload intensity. Each of those figures show the speedup with
different dataset sizes (e.g., $838,000$, $2,567,600$ atoms, etc.),
including $10$, $100$, and $1000$ data frames. The speedup is
calculated simply as a ratio between the running time of our system
on a certain set of workload and that of the Gromacs system on the
same workload. These figures show that the speedup over varying
workload intensity achieved by our system ranges anywhere from about
$10$ to $1000$ times, depending on the size of the dataset, number
of frames and the selection of the atoms.

Figure~\ref{fg:speedup-average-over-all-workload}, shows the speedup
our system achieves over all workload intensity (average workload
intensity) with varying dataset sizes. It is clear that, again our
system has better performance than the Gromacs system. The speedup
presented in this set of figures ranges anywhere from about $15$ to
$650$ times.

The last figure, Figure~\ref{fg:speedup-average-workload-select},
shows the speedup our system achieves over all workload intensity
and all select levels with varying dataset sizes. This figure, in a
way, summarizes the previous two sets of figures, bringing together
the workload and the different selections through the average. It is
clear that, again our system has better performance than the Gromacs
system. The speedup ranges anywhere from $50$ to $250$.

All four sets of figures show that such push-based design has clear
advantages over the pull-based type of design incorporated in the
Gromacs system.

@@

\noindent{\bf Acknowledgements:} The project described was supported
by an Award (R01GM086707) from the National Institute Of General
Medical Sciences (NIGMS) at the National Institutes of Health (NIH).
The authors would like to thank Anand Kumar who has contributed his
time and knowledge toward completion of the work.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\emph{\textbf{SDH implementation.}} Even though implementation of SDH can be different, in the previous paper the algorithm in the previous paper was based on quad tree data structure with data proximity approach based on data uniformity\cite{mainPaper}, but in our experiments we follow special CUDA based approach which perfectly fits the nature of GPU devices with the given memory hierarchy, and the number of computations are equal to brute force, consequently, there is no expectation on data uniformity, thus, the problem solution complexity is still the same with the worst case of $O(N^2)$, but processed in parallel by GPU cores. More improvements can be done for acceleration of this approach, but in the system we have not gone too far with data converter module, because initial framework has to be simple and user friendly.

\subsection{Results}
Results are nicely represented, in general, on Figures \ref{fg:chart_1000}, \ref{fg:chart_10000} \ref{fg:chart_max}. You can instantly notice that speedup for 1000 atoms is not very impressive, especially for $10$ frames, and the reason for that is the following. There is a small overhead we have to consider before computing part. In order to let GPU process the data, we have to transfer the data to Global Memory, and upon finish we have to load the results back to RAM. In this particular case, for small workload and frames, it is even slower than the sequential processing. Afterwards, the bigger workload the more convincing speedup gets, unfortunately, it doesn't get more than $4$ times for 1000 atoms. Fortunately, scientists usually work with much larger data sets, thus, this doesn't really represent general speedup for average use case.

\emph{\textbf{10000 atoms.}}
As you can see, speedup bumps up to almost 30 with 10000 atoms (Figure \ref{fg:chart_10000}). Once the overhead on memory transfer is covered by computation speedup, it keeps growing. It is important to pay attention to the growth of speedup over growth of workload. It is not that significant, because even though CPUs are much slower, they still use smart caching and internal multi-core processing, which is pretty fast, too.

@@

@@
